Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-06-01 docs/submission/response_to_reviewer_comments/reviewer_1/response_reviewer_1.tex

\noindent  We hope that this satisfies the request for changes necessary to proceed with the publication of the updated manuscript. \newline

\section*{Essential comments}

\begin{enumerate}
	\item \textbf{Essential comment 1: mixing of different models, parameters and predictors}

	We agree with the general points here raised by the reviewer that the terminology used was insufficiently precise. This has now been corrected throughout the entire manuscript. We now make it clear that we are always comparing the results between two neural network models, not comparing NN predictions with e.g. FLake. If we update some input field to our NN and the NN prediction accuracy improves, this is good evidence that the updated field itself is more accurate and would enable a physical model like FLake to make more accurate predictions. We note that this work is a first attempt / investigation to explore the possibility of using these kind of machine learning methods for a global check of surface physiographic fields, and we will look to refine and further develop these methods in future.

@@

	Taking some individual points:

	\begin{itemize}
		\item \textbf{L.79, L82-87}. What was meant here is the distinction between is the new information informative enough to have visible changes, or it is just lots of work for no impact? This has now been properly phrased in the text. Whilst we did briefly discuss local degradation previously we have also now included a more explicit discussion. Additionally, we explicitly calculate the training noise by retraining VESPER multiple times. We now highlight the effect of the training noise in the updated manuscript. Having retrained the model multiple times, our conclusions are generally unchanged for all grid-point categories, with the exception of the Vegetation category which has significant training noise over a small number of grid cells making it difficult to draw meaningful conclusions. This is again discussed in the text and we thank the reviewer for raising this point for our attention. 		
		\item \textbf{2.2 L88-89} As the point above, we have retrained VESPER multiple times to show that the training noise is generally smaller than the prediction changes due to the different input fields. We reword the text throughout to be clear that we are always comparing two NN models.
		\item \textbf{2.3 L105-106}  As above, terminology and text has been changed throughout the manuscript to be more precise
		\item \textbf{2.4 L19}  As discussed we have now trained multiple versions of the model to better quantify the training noise and our conclusions remain unchanged that we can use VESPER to (a) check that an updated field is closer to reality and (b) see if this updated field increases the accuracy of our NN model. Both of these points are relevant for the updated fields within a physical model like FLake.
	\end{itemize}

		\item \textbf{Essential comment 2: strange results and speculative explanations}

		As mentioned we now explicitly calculate the training noise by retraining VESPER multiple times. For the points in Northern Canada, the Toshka lakes and the Vegetation category our previously quoted changes are less than the training noise. Again we thank the reviewer for highlighting this to us. We have removed the discussion on  Northern Canada and the Toshka lakes from the manuscript, whilst the effect of the noise on the Vegetation category is now discussed explicitly in the text. We emphasize that our main conclusion re the lake and glacier categories are unchanged by this retraining - the difference in the improvement due to the updated fields is much greater than the training noise.

### 2023-06-01 docs/submission/template.tex
final corrections

\begin{abstract}
A purposely built deep learning algorithm for the Verification of Earth-System ParametERisation (VESPER) is used to assess recent upgrades of the global physiographic datasets underpinning the quality of the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF), which is used both in numerical weather prediction and climate reanalyses. A neural network regression model is trained to learn the mapping between the surface physiographic dataset plus the meteorology from ERA5, and the MODIS satellite skin temperature observations. Once trained, this tool is applied to rapidly assess the quality of upgrades of the land-surface scheme. Upgrades which improve the prediction accuracy of the machine learning tool indicate a reduction of the errors in the surface fields used as input to the surface parametrisation schemes. Conversely, incorrect specifications of the surface fields decrease the accuracy with which VESPER can make predictions.  We apply VESPER to assess the accuracy of recent upgrades of the permanent lake and glaciers covers as well as planned upgrades to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature improves by 0.37 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 0.84 K. We also show that updates to the glacier cover improve further the prediction accuracy by 0.22 K. We highlight how neural networks such as VESPER can assist the research and development of surface parametrizations and their input physiography to better represent Earth’s surface coupled processes in weather and climate models.

@@

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
Accurate knowledge of the global surface physiography, including land, water and ice covers, and their characteristics, strongly determines the quality of surface and near-surface temperature simulations in weather and climate modelling. For instance, water bodies exchange mass and energy with the atmosphere and their thermal inertia strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic, over saturated with CO$_2$, as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO2-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Huang2019,Peng2020,Franz2018}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent Lakes are incorporated in Earth-system models via parametrisation. At ECMWF the  representation of lakes via parametrisation was first handled by introducing the Fresh water Lake model FLake \citep{Mironov2008} into the IFS. FLake treats all resolved inland waterbodies (i.e. lakes, reservoirs, rivers which are dominating in a grid-cell) and unresolved or sub-grid water (i.e. small inland waterbodies and sea/ocean coastal waters which are present but not dominating in a grid-cell). Note that lake parameters are also an important part of the FLake model so when we refer in this work to ``lake parametrisation" we mean both the model and the parameters. The broad impact of the FLake model and the important role that waterbodies play in human life can be illustrated by analysing ECMWF maps of the fractional land sea mask and the inland waterbody cover alongside maps of the population density (i.e. inhabitants per km$^2$) based on the population count for 2015 from the Global Human Settlement Layers (GHSL), Population Grid 1975-2030 \citep{GHS,JRC100523} at 9 km horizontal resolution. Globally FLake is active over 11.1\% of the grid-cells, with only 1.2\% of them being resolved inland waters (i.e. water covers $\geq$50\% of the grid-cell); considering only non-ocean (i.e. land) grid-cells, then FLake is active over 32.4\% of the grid-cells with only 3.5\% of them being resolved waters. According to the population data, only 4\% of land is densely populated (i.e. at least 300 inhabitants per km$^2$); 64.5\% of these areas being situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast) with half of it (i.e. 31.2\% of densely populated areas) being in the vicinity of at least 1 km$^2$ waterbody - emphasising how essential waterbodies are in human life. In some regions this role may be even more crucial than in the others. For example, only 2\% of the North American region (similar for South American and North Asian regions) is densely populated with 45.7\% (33.9\% and 37.9\% respectively) of the areas being in vicinity of at least 1 km$^2$ waterbody; for Europe even though it has more densely populated areas (16\% of land is densely populated) still 37.4\% of the population are in the vicinity of at least a 1 km$^2$ waterbody; for a rather dry continent like Africa only 5\% of land is densely populated with 22.2\% of these areas being close to at least a 1 km$^2$ waterbody; most striking in this sense is Australia where only 0.5 \% of the land is populated, with two thirds of the population living within 9 km radius of a permanent waterbody of at least 1 km$^2$, with the majority of people living on the ocean coast. \newline

@@

		\\
		Additional surface physiographic fields:               & Difference for all main surface physiographic fields between V15 and V20 field sets, \newline 
		Difference between V20 static lake cover and monthly varying lake cover (12 maps in total),\newline 
		Saline lake cover (the proportion of a grid-cell covered by saline inland water bodies; units: 0-1)
		 \\
		\hline
	\end{tabularx}
\caption{Input features used for training the neural network model VESPER; atmospheric model fields (time varying) were kept the same in all simulations,  surface  physiographic  fields  (static)  were  updated  when  going  from  the  original  data  based  on GlobeCover2009/GLDBv1 (V15 field set) to GSWE/GLDBv3 (V20 field set); in brackets are variables description (where needed), short name (according to the GRIB parameter database) and units.
}
\label{table:definitions}
\end{table*}

@@

\noindent Aqua  \citep{aquaref} is a NASA satellite mission which makes up part of the Earth Observing System (EOS). Operating at an altitude of 700km, with orbital period of 99 minutes, its orbital trajectory passes south to north with an equatorial-crossing times in general of 1.30pm. This post-meridian crossing time has led to it sometimes being denoted as EOS PM. Launched in 2002 with an initial expected mission duration of 6 years, Aqua has far exceeded its initial brief and continues to transmit information from 4 of the 6 observation instruments on board. Here we use information only from MODIS instrument. MODIS can take surface temperature measurements at a spatial resolution of 1km (the exact grid size is 0.928km by 0.928km), operating in the wavelength ranges of between ∼3.7-4.5µm and ∼10.9-12.3µm. In addition to surface temperature measurements that were used in this work, MODIS can take observations of cloud properties, water vapour, ozone etc. Here MYD11A1 v006 \citep{MODISusersguide} collection that provides daily LST measurements at a spatial resolution of 1km on a sinusoidal projection grid SR-ORG:6974 (takes a spherical projection but a WGS84 datum ellipsoid) is exercised. Daily global LST data is generated by first applying a split-window LST algorithm \citep{508406} on all nominal (i.e. 1km at nadir) resolution swath (scene) with a nominal coverage of ~5 minutes of MODIS scans along the track acquired in daytime, and secondly by mapping results onto integerized sinusoidal projection; for more details see \citep{MODISusersguide}. Validation of this product was carried out using temperature-based method over different land cover types (e.g. grasslands, croplands, shrublands, woody areas, etc.) in several regions around the globe (i.e. United States, Portugal, Namibia, and China) at different atmospheric and/or surface conditions; the best accuracy is achieved over United States sites with RMSE lower than 1.3K \citep{DUAN201916}. At large view angles and in semi-arid regions the data product may have slightly higher errors due to rather uncertain classification-based surface emissivities and heavy dust aerosols effects. In addition, the MODIS cloud mask struggles to eliminate all cloud and/or heavy aerosols contaminated grid-cells from the clear-sky ones (LST errors in such grid-cells can be 4-11K and larger). Validation of this product over five bare ground sites in north Africa (in total 12 radiosonde-based datasets validated) showed that mean LST error was within $\pm$0.6K (with exception for one dataset, where mean LST error was 0.8K) and standard deviation of LST errors were less than 0.5K \cite{DUAN201916}. In this work to reduce the amount of daily data over multiple years to store and manipulate, prior use LST data is (i) filtered to contain only cloud free data, and (ii) averaged to a ~4km at the Equator resolution on a regular latitude-longitude grid, EPSG 4326 (note that only grid cells which have 8 or more valid observations at 1km resolution are averaged over, otherwise they are classified as missing data).

%\begin{figure}
%	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{example-image-c}} \\
%	\caption{A brief step-by-step explanation of the LST algorithm for MYD11A1 v006 collection.} 
%	\label{fig:example_figure_c}
%\end{figure}

\subsection{Joining the data}\label{sec:join}
To join selected ERA5 global fields on a reduced Gaussian grid at ~31km resolution (information in UTC, 24 hourly maps per day) with Aqua-MODIS global LST data on a regular latitude-longitude grid at ~4km resolution (information in local solar time, 1 map per day), both datasets need to be at the same time space. First it is necessary to determine the absolute time (i.e. UTC) at which the MODIS observations were taken. Since in general all Aqua observations are taken at 1.30pm local solar time, it can be related to a UTC via observation longitude, following Eq. \ref{eq:time_conversion}:
\begin{equation}
	\text{UTC} = \text{Local solar time} - \frac{\text{longitude}}{15} \ , 
	\label{eq:time_conversion}
\end{equation}
where longitude is in degrees, and UTC is rounded to the nearest hour. This conversion is inexact since there is an additional correction as a function of the latitude, yet is the de factor recommended by the official MODIS Products User’s Guide \cite{MODISusersguide}; given the short orbital period of Aqua these additional higher order corrections are expected to be typically small and for our purposes can be neglected. Also, the assumption that all Aqua observations are taken at 1.30pm local solar time was checked (see Figure \ref{fig:MODIS_time_error}). The annually averaged mean time difference at ~31km resolution (i.e. daily differences between local solar time of observations and 1.30pm at 1km resolution were first aggregated to 31km resolution using averaging, and then aggregated in time over a year) is 0.16 hours or 10 min, with mean absolute error (MAE) being 0.46 hours or 28 min and root mean square error (RMSE) being 0.61 hours or 37 min (current values correspond 70N-70S region year 2019, but confirmed to be approximately identical for each year of 2016-2019 period). Since the temporal resolution of ERA5 data is hourly, the assumptions inherent to Eq \ref{eq:time_conversion} are sufficiently accurate. Over the poles (i.e. 90-70°N and 70-90°S) satellite sweeps overlap significantly and in general conversion becomes less accurate (daily time differences can reach more than $\pm$ 3.5 hours), so these areas were not included in the analysis. \newline

@@

\begin{figure}
	\subfloat[\label{fig:modis_plot_1}]{\includegraphics[width=0.48\textwidth]{num_obs_map}} \\
	\subfloat[\label{fig:modis_plot_2}]{\includegraphics[width=0.48\textwidth]{MODIS_obs_error}}
	\caption{For 2019 at ~31km resolution: (a) Mean daily number of Aqua-MODIS observations mapped to each ERA5 data point. The swath of the Aqua satellite is clearly visible, with more observations over 70-60°N and 60-70°S areas as Aqua follows a polar orbit, south to north, and with less observations over Equator, complex orography areas (such as the Himalayas, the Andes and the Rocky Mountains), and the Siberian Tundra (due to increased cloud cover). (b) Average error in the Aqua-MODIS LST measurement. The raw Aqua-MODIS data at ~1km resolution provides categorical LST errors with bins ≤ 1K, 1-2K, 2-3K and > 3K. When averaging to the coarser resolution a weighted average over the 1km grid-cells is computed, where the median bin value is used, and 5K for the > 3K bin. This information helps to understand that abundant number of observation does not automatically mean high quality of LST (e.g. Australia).} 
	\label{fig:MODIS_time_error_N}
\end{figure}
\noindent For step (iii) in the joining process, The second step we use a GPU-accelerated k-nearest neighbours algorithm \cite{Rapids}, where “nearness” on the sphere between two points is measured via the Haversine metric i.e. the geodesic distance H, following Eq. (2):
\begin{equation}
	H = 2 \arcsin (d)
\end{equation}
where
\begin{eqnarray}
	d = \sqrt{\sin^2\left(\frac{\delta\theta}{2}\right) + \cos \theta_1 \cos \theta_2 \sin^2\left(\frac{\delta \phi}{2}\right) }
\end{eqnarray} 
for two points with coordinate latitudes $\theta_{1,2}$, longitudes $\phi_{1,2}$ and $\delta \theta = \theta_2 - \theta_1$ and  $\delta \phi = \phi_2 - \phi_1$ .

@@

 \noindent VESPER was trained on selected atmospheric and surface model fields from ERA5 (see Table \ref{table:definitions}), certain static version of the surface physiographic fields (see Table \ref{tab:datasources}), and Aqua-MODIS LST for 2016. Once VESPER was fully trained it was used to predict LST over the whole globe for 2019.  Going forward, as a shorthand we will refer to VESPER trained using $VM$ surface physiographic fields as VESPER\_VM. For instance, a VESPER model trained using the surface physiographic fields from V15, we refer to as VESPER\_V15 (see Table \ref{tab:vesper_table} for an explicit definition of all the VESPER models). All VESPER versions are trained with ERA5 fields for 2016 and with main surface physiographic fields V15. Then depending on the version some or all additional surface physiographic fields (see Table \ref{table:definitions}) are added. VESPER’s predictions can be compared to the initial ERA5 skin temperatures and actual Aqua-MODIS LST for 2019. Figure \ref{fig:example_model} shows the mean absolute errors (MAE) globally in the VESPER\_V15 LST predictions, relative to the Aqua-MODIS LST along with the corresponding MAE in the predicted skin temperature from ERA5. We can see that VESPER\_V15 was able to learn corrections to ERA5, especially in the Himalayas and sub-Saharan Africa as well as Australia and the Amazon basin, leading to the globally averaged MAE reduction for predicted LST; the MAE relative to  Aqua-MODIS LST, averaged over all grid points, was 3.9K for ERA5 and 3.0K for VESPER\_V15. \newline

>>>>

\begin{abstract}
A purposely built deep learning algorithm for the Verification of Earth-System ParametERisation (VESPER) is used to assess recent upgrades of the global physiographic datasets underpinning the quality of the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF), which is used both in numerical weather prediction and climate reanalyses. A neural network regression model is trained to learn the mapping between the surface physiographic dataset plus the meteorology from ERA5, and the MODIS satellite skin temperature observations. Once trained, this tool is applied to rapidly assess the quality of upgrades of the land-surface scheme. Upgrades which improve the prediction accuracy of the machine learning tool indicate a reduction of the errors in the surface fields used as input to the surface parametrisation schemes. Conversely, incorrect specifications of the surface fields decrease the accuracy with which VESPER can make predictions.  We apply VESPER to assess the accuracy of recent upgrades of the permanent lake and glaciers covers as well as planned upgrades to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature (i.e mean absolute error difference between updated and original physiographic datasets) improves by 0.37 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 0.83 K. We also show that updates to the glacier cover improve the prediction accuracy by 0.22 K. We highlight how neural networks such as VESPER can assist the research and development of surface parametrizations and their input physiography to better represent Earth’s surface coupled processes in weather and climate models.

@@

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
Accurate knowledge of the global surface physiography, including land, water and ice covers, and their characteristics, strongly determines the quality of surface and near-surface temperature simulations in weather and climate modelling. For instance, water bodies exchange mass and energy with the atmosphere and their thermal inertia strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic (i.e. over saturated with carbon dioxide, CO$_2$), as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO2-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication (i.e. nutrient concentration increase) of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Huang2019,Peng2020,Franz2018}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent Lakes are incorporated in Earth-system models via parametrisation. At ECMWF the  representation of lakes via parametrisation was first handled by introducing the Fresh water Lake model FLake \citep{Mironov2008} into the IFS. FLake treats all resolved inland waterbodies (i.e. lakes, reservoirs, rivers which are dominating in a grid-cell) and unresolved or sub-grid water (i.e. small inland waterbodies and sea/ocean coastal waters which are present but not dominating in a grid-cell). Note that lake parameters are also an important part of the FLake model so when we refer in this work to ``lake parametrisation" we mean both the model and the parameters. The broad impact of the FLake model and the important role that waterbodies play in human life can be illustrated by analysing ECMWF maps of the fractional land sea mask and the inland waterbody cover alongside maps of the population density (i.e. inhabitants per km$^2$) based on the population count for 2015 from the Global Human Settlement Layers (GHSL), Population Grid 1975-2030 \citep{GHS,JRC100523} at 9 km horizontal resolution.

\hl{Globally FLake is active over 11.1\% of the grid-cells; considering only land grid-cells, then FLake is active over 32.4\% of the points. According to the population data, 64.5\% of densely populated areas (at least 300 inhabitants per km$^2$) are situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast), with 31.2\% being in the vicinity of at least 1 km$^2$ waterbody - emphasising how essential waterbodies are in human life. In some regions this role may be even more crucial than in the others. For example in North America 45.7\% of the densely populated areas are close to a 1 km$^2$ waterbody; in Australia where only 0.5 \% of the land is populated,  two thirds of the population live within 9 km radius of a permanent waterbody of at least 1 km$^2$, with the majority of people living on the ocean coast.} \newline

%\hl{Globally FLake is active over 11.1\% of the grid-cells, with only 1.2\% of them being resolved inland waters (i.e. water covers $\geq$50\% of the grid-cell); considering only non-ocean (i.e. land) grid-cells, then FLake is active over 32.4\% of the grid-cells with only 3.5\% of them being resolved waters. According to the population data, only 4\% of land is densely populated (i.e. at least 300 inhabitants per km$^2$); 64.5\% of these areas being situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast) with half of it (i.e. 31.2\% of densely populated areas) being in the vicinity of at least 1 km$^2$ waterbody - emphasising how essential waterbodies are in human life. In some regions this role may be even more crucial than in the others. For example, only 2\% of the North American region (similar for South American and North Asian regions) is densely populated with 45.7\% (33.9\% and 37.9\% respectively) of the areas being in vicinity of at least 1 km$^2$ waterbody; for Europe even though it has more densely populated areas (16\% of land is densely populated) still 37.4\% of the population are in the vicinity of at least a 1 km$^2$ waterbody; for a rather dry continent like Africa only 5\% of land is densely populated with 22.2\% of these areas being close to at least a 1 km$^2$ waterbody; most striking in this sense is Australia where only 0.5 \% of the land is populated, with two thirds of the population living within 9 km radius of a permanent waterbody of at least 1 km$^2$, with the majority of people living on the ocean coast.} \newline

@@

		\\
		Additional surface physiographic fields               & Difference for all main surface physiographic fields between V15 and V20 field sets, \newline 
		Difference between V20 static lake cover and monthly varying lake cover (12 maps in total),\newline 
		Saline lake cover (the proportion of a grid-cell covered by saline inland water bodies; units: 0-1)
		 \\
		\hline
	\end{tabularx}
\caption{Input features used for training the neural network model VESPER; atmospheric model fields (time varying) were kept the same in all simulations,  surface  physiographic  fields  (static)  were  updated  when  going  from  the  original  data  based  on GlobeCover2009/GLDBv1 (V15 field set) to GSWE/GLDBv3 (V20 field set); in brackets are variables description (where needed), short name (according to the GRIB parameter database) and units.
}
\label{table:definitions}
\end{table*}

@@

\noindent Aqua  \citep{aquaref} is a NASA satellite mission which makes up part of the Earth Observing System (EOS). Operating at an altitude of 700km, with orbital period of 99 minutes, its orbital trajectory passes south to north with an equatorial-crossing times in general of 1.30pm. This post-meridian crossing time has led to it sometimes being denoted as EOS PM. Launched in 2002 with an initial expected mission duration of 6 years, Aqua has far exceeded its initial brief and continues to transmit information from 4 of the 6 observation instruments on board. Here we use information only from MODIS instrument. MODIS can take surface temperature measurements at a spatial resolution of 1km (the exact grid size is 0.928km by 0.928km), operating in the wavelength ranges of between ∼3.7-4.5µm and ∼10.9-12.3µm. In addition to surface temperature measurements that were used in this work, MODIS can take observations of cloud properties, water vapour, ozone etc. Here MYD11A1 v006 \citep{MODISusersguide} collection that provides daily LST measurements at a spatial resolution of 1km on a sinusoidal projection grid SR-ORG:6974 (takes a spherical projection but a WGS84 datum ellipsoid) is exercised. Daily global LST data is generated by first applying a split-window LST algorithm \citep{508406} on all nominal (i.e. 1km at nadir) resolution swath (scene) with a nominal coverage of ~5 minutes of MODIS scans along the track acquired in daytime, and secondly by mapping results onto integerized sinusoidal projection; for more details see \citep{MODISusersguide} and Figure \ref{fig:example_figure_c}. Validation of this product was carried out using temperature-based method over different land cover types (e.g. grasslands, croplands, shrublands, woody areas, etc.) in several regions around the globe (i.e. United States, Portugal, Namibia, and China) at different atmospheric and/or surface conditions; the best accuracy is achieved over United States sites with RMSE lower than 1.3K \citep{DUAN201916}. At large view angles and in semi-arid regions the data product may have slightly higher errors due to rather uncertain classification-based surface emissivities and heavy dust aerosols effects. In addition, the MODIS cloud mask struggles to eliminate all cloud and/or heavy aerosols contaminated grid-cells from the clear-sky ones (LST errors in such grid-cells can be 4-11K and larger). Validation of this product over five bare ground sites in north Africa (in total 12 radiosonde-based datasets validated) showed that mean LST error was within $\pm$0.6K (with exception for one dataset, where mean LST error was 0.8K) and standard deviation of LST errors were less than 0.5K \cite{DUAN201916}. In this work to reduce the amount of daily data over multiple years to store and manipulate, prior use LST data is (i) filtered to contain only cloud free data, and (ii) averaged to a ~4km at the Equator resolution on a regular latitude-longitude grid, EPSG 4326 (note that only grid cells which have 8 or more valid observations at 1km resolution are averaged over, otherwise they are classified as missing data).

\begin{figure}
	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{MODIS_flow}} \\
	\caption{A brief step-by-step explanation of the LST algorithm for MYD11A1 v006 collection.} 
	\label{fig:example_figure_c}
\end{figure}

\subsection{Joining the data}\label{sec:join}
To join selected ERA5 global fields on a reduced Gaussian grid at $\sim$ 31km resolution (information in UTC, 24 hourly maps per day) with Aqua-MODIS global LST data on a regular latitude-longitude grid at ~4km resolution (information in local solar time, 1 map per day), both datasets need to be at the same time space. First it is necessary to determine the absolute time (i.e. UTC) at which the MODIS observations were taken. Since in general all Aqua observations are taken at 1.30pm local solar time, it can be related to a UTC via observation longitude, following Eq. \ref{eq:time_conversion}:
\begin{equation}
	\text{UTC} = \text{Local solar time} - \frac{\text{longitude}}{15} \ , 
	\label{eq:time_conversion}
\end{equation}
where longitude is in degrees, and UTC is rounded to the nearest hour. This conversion is inexact since there is an additional correction as a function of the latitude, yet is the de factor recommended by the official MODIS Products User’s Guide \cite{MODISusersguide}; given the short orbital period of Aqua these additional higher order corrections are expected to be typically small and for our purposes can be neglected. Also, the assumption that all Aqua observations are taken at 1.30pm local solar time was checked (see Figure \ref{fig:MODIS_time_error}). The annually averaged mean time difference at ~31km resolution (i.e. daily differences between local solar time of observations and 1.30pm at 1km resolution were first aggregated to 31km resolution using averaging, and then aggregated in time over a year) is 0.16 hours or 10 min, with mean absolute error (MAE) being 0.46 hours or 28 min and root mean square error (RMSE) being 0.61 hours or 37 min (current values correspond 70N-70S region year 2019, but confirmed to be approximately identical for each year of 2016-2019 period). Since the temporal resolution of ERA5 data is hourly, the assumptions inherent to Eq \ref{eq:time_conversion} are sufficiently accurate. Over the poles (i.e. 90-70°N and 70-90°S) satellite sweeps overlap significantly and in general conversion becomes less accurate (daily time differences can reach more than $\pm$ 3.5 hours), so these areas were not included in the analysis. \newline

@@

\begin{figure}
	\subfloat[\label{fig:modis_plot_1}]{\includegraphics[width=0.48\textwidth]{num_obs_map}} \\
	\subfloat[\label{fig:modis_plot_2}]{\includegraphics[width=0.48\textwidth]{MODIS_obs_error}}
	\caption{For 2019 at ~31km resolution: (a) Mean daily number of Aqua-MODIS observations mapped to each ERA5 data point. The swath of the Aqua satellite is clearly visible, with more observations over 70-60°N and 60-70°S areas as Aqua follows a polar orbit, south to north, and with less observations over Equator, complex orography areas (such as the Himalayas, the Andes and the Rocky Mountains), and the Siberian Tundra (due to increased cloud cover). (b) Average error in the Aqua-MODIS LST measurement. The raw Aqua-MODIS data at ~1km resolution provides categorical LST errors with bins ≤ 1K, 1-2K, 2-3K and > 3K. When averaging to the coarser resolution a weighted average over the 1km grid-cells is computed, where the median bin value is used, and 5K for the > 3K bin. This information helps to understand that abundant number of observation does not automatically mean high quality of LST (e.g. Australia).} 
	\label{fig:MODIS_time_error_N}
\end{figure}
\noindent For step (iii) in the joining process, The second step we use a GPU-accelerated k-nearest neighbours algorithm \cite{Rapids}, where “nearness” on the sphere between two points is measured via the Haversine metric i.e. the geodesic distance $H$, following Eq. (2):
\begin{equation}
	H = 2 \arcsin \left(\sqrt{\sin^2\left(\frac{\delta\theta}{2}\right) + \cos \theta_1 \cos \theta_2 \sin^2\left(\frac{\delta \phi}{2}\right) }\right)
\end{equation}
for two points with coordinate latitudes $\theta_{1,2}$, longitudes $\phi_{1,2}$ and $\delta \theta = \theta_2 - \theta_1$ and  $\delta \phi = \phi_2 - \phi_1$ .

%where
%\begin{eqnarray}
%	d = \sqrt{\sin^2\left(\frac{\delta\theta}{2}\right) + \cos \theta_1 \cos \theta_2 \sin^2\left(\frac{\delta \phi}{2}\right) }
%\end{eqnarray}

@@

 \noindent VESPER was trained on selected atmospheric and surface model fields from ERA5 (see Table \ref{table:definitions}), certain static version of the surface physiographic fields (see Table \ref{tab:datasources}), and Aqua-MODIS LST for 2016. Once VESPER was fully trained it was used to predict LST over the whole globe for 2019. \hl{The training and test years were chosen simply as recent, non-anomalous years where the updated  surface physiographic fields should provide more accurate information.} Going forward, as a shorthand we will refer to VESPER trained using $VM$ surface physiographic fields as VESPER\_VM. For instance, a VESPER model trained using the surface physiographic fields from V15, we refer to as VESPER\_V15 (see Table \ref{tab:vesper_table} for an explicit definition of all the VESPER models). All VESPER versions are trained with ERA5 fields for 2016 and with main surface physiographic fields V15. Then depending on the version some or all additional surface physiographic fields (see Table \ref{table:definitions}) are added. VESPER’s predictions can be compared to the initial ERA5 skin temperatures and actual Aqua-MODIS LST for 2019. Figure \ref{fig:example_model} shows the mean absolute errors (MAE) globally in the VESPER\_V15 LST predictions, relative to the Aqua-MODIS LST along with the corresponding MAE in the predicted skin temperature from ERA5. We can see that VESPER\_V15 was able to learn corrections to ERA5, especially in the Himalayas and sub-Saharan Africa as well as Australia and the Amazon basin, leading to the globally averaged MAE reduction for predicted LST; the MAE relative to  Aqua-MODIS LST, averaged over all grid points, was 3.9K for ERA5 and 3.0K for VESPER\_V15. \newline
